---
title: "Assignment_1_NDVI_Trace"
author: "Nicole Ross"
date: "9/9/2019"
output: html_document
# bibliography: cite.bib
---

```{r setup, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidyr)
library(ggthemes)
library(lubridate)
```

## Assignment 1

This markdown file works on assignment 1 in Hayman Fire NDVI Trace.


### Reading in data

This code takes in the separate NDMI, NDSI, and NDVI datasets, cleans them, and then stacks them into one dataset called "full_long".

```{r data_read, warning = FALSE, message = FALSE}
files <- list.files('data',full.names=T)
files
#View(files)

#Read in individual data files
ndmi <- read.csv(files[1]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndmi')

ndsi <- read.csv(files[2]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndsi')

ndvi <- read_csv(files[3])%>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndvi')

```


### Stacking and Gathering the Data

I tried to test my coding knowledge of R by attempting to get rid of NAs *before* gathering. However, this approach does not work because I get rid of some observations that do not have NAs in them.

Remember, that *arrange* is the equivalent of *sort* or *gsort* in Stata, while *gather* 
is the equivalent of *reshape wide to long* in Stata, and *spread* is the equivalent of 
*reshape long to wide* in Stata.

```{r stacking_gathering}
 
# Stack as a tidy dataset
  full_long <- rbind(ndvi,ndmi,ndsi) %>%
  arrange(DateTime, data) %>%
  
#Gather the data  
gather(key='site',value='value',-DateTime,-data) %>%
  filter(!is.na(value)) %>%
  arrange(DateTime, site)
  #View(full_long)
```


### Question 1

What is the correlation between NDVI and NDMI? - here I want you to convert the full_long dataset in to a wide dataset using the function "spread" and then make a plot that shows the correlation as a
function of if the site was burned or not


#### Nicole's Code and Comments

For ease, I named the plot here, p1. To show correlations and perform other related analyses (e.g., regressions, etc.), I need to *spread* the data into a wide dataset.
  

```{r question 1}

#View(full_long)
?spread
full_wide <- spread(data=full_long, key='data', value='value') %>%
  filter_if(is.numeric, all_vars(!is.na(.)))
    #summary(full_wide)
    #View(full_wide)

    p1 <- ggplot(full_wide, aes(x=ndmi, y=ndvi, color=site)) +
        geom_point() +
        theme_few() +
        scale_color_few() +
        theme(legend.position=c(.2, .2))
    p1
```


### Question 2

What is the correlation between average NDSI (normalized snow index) for January - April and average NDVI for June-August? In other words, does the previous year's snow cover influence vegetation
growth for the following summer? 


#### Nicole's Code and Comments

For this question, I will need to use the *filter* command, which is the equivalent of selecting rows of data that match the criteria I give it to filter through. I will also need to *mutate* the data, which is the equivalent of *generate* in Stata. I will also need to use the *group_by* command.

This question also entails making two separate datasets and then *merging* them back together again (Stata speak), which is the equivlanet of using a *join* in R.

```{r question 2}

full_wide <- spread(data=full_long, key='data', value='value') %>%
    filter_if(is.numeric, all_vars(!is.na(.))) %>%
    mutate(month = month(DateTime),
          year = year(DateTime)) 
    
winter_only <- filter(full_wide, month %in% c(1,2,3,4)) %>%
    group_by(year) %>%
    summarize(ndsi=mean(ndsi))

summer_only <- filter(full_wide, month %in% c(6,7,8)) %>%
    group_by(year) %>%
    summarize(ndvi=mean(ndvi))

winter_summer_together <- inner_join(winter_only, summer_only, by = "year")

      p2 <- ggplot(winter_summer_together, aes(x=ndsi, y=ndvi)) +
          geom_point() +
          theme_few() +
          scale_color_few() +
          theme(legend.position=c(.2, .2))
      p2
```

Based on the plot, there does not appear to be a correlation between NDSI and NDVI.


### Question 3

How is the snow effect from question 2 different between pre- and post-burn and burned and unburned? 


#### Nicole's Code and Comments

This question appears to be asking us to replicate question 2 plot, but splitting the points along the two dimensions. I make three plots below to show the relationship depending on how the question is worded.

```{r question 3}

full_wide_treatment <- spread(data=full_long, key='data', value='value') %>%
    filter_if(is.numeric, all_vars(!is.na(.))) %>%
    mutate(month = month(DateTime),
         year = year(DateTime),
         treatment = cut(year,breaks=c(0,2003,2020),
                labels=c('pre-burn','post-burn'))) 
  
 winter_only_treatment <- filter(full_wide_treatment, month %in% c(1,2,3,4)) %>%
      group_by(year, site, treatment) %>%
      summarize(ndsi=mean(ndsi))
    
    summer_only_treatment <- filter(full_wide_treatment, month %in% c(6,7,8)) %>%
      group_by(year, site, treatment) %>%
      summarize(ndvi=mean(ndvi))
    
    winter_summer_treatment_together <- inner_join(winter_only_treatment, 
      summer_only_treatment, by = c("year" = "year", "site" = "site", "treatment"= "treatment"))
     
     
    
    p3 <- ggplot(winter_summer_treatment_together, aes(x=ndsi, y=ndvi, color=treatment)) +
      geom_point() +
      theme_few() +
      scale_color_few() +
      theme(legend.position=c(.8, .2))
    p3
    
    
    p4 <- ggplot(winter_summer_treatment_together, aes(x=ndsi, y=ndvi, color=site)) +
      geom_point() +
      theme_few() +
      scale_color_few() +
      theme(legend.position=c(.8, .2))
    p4
    
    
     p5 <- ggplot(winter_summer_treatment_together, aes(x=ndsi, y=ndvi, color=site)) +
      geom_point() +
      theme_few() +
      scale_color_few() +
      theme(legend.position=c(.2, .2)) +
      facet_wrap(~treatment)
    p5

```

The points are clustered in the post-burn period. Pre-burn, there is no relationship between NDSI and NDVI in either the burned or unburned areas. Howeer, in the post-burn era, the same value of NDSI yields differnt NDVI values for the burned and unburned areas. Burned areas, on average, had a lower NDVI for the same average NDSI value compared to unburned areas.


### Question 4 

What month is the greenest month on average? Does this change in the burned plots after the fire? 


#### Nicole's Code and Comments

This question requires us to aggregate information to each month and then plot that.

```{r question 4.1}

ndvi_month_wide <-select(full_wide, -ndmi, -ndsi) %>%
                  group_by(month) %>%
                  summarize(ndvi=mean(ndvi))

#View(ndvi_month_wide)

  p6 <- ggplot(ndvi_month_wide, aes(x=month, y=ndvi)) +
      geom_point() +
      theme_few() +
      scale_color_few()
  p6
    
```

According to the NDVI data, the greenest month on average, inclusive of years 1984 to present, is August.

If we want to know how this changed in the burned plots after the fire, I will want to plot NDVI pre-burn and post-burn by month.

```{r question 4.2}

ndvi_month_wide_burn <-select(full_wide, site, ndvi, month, year) %>%
                  filter(site=="burned") %>%
                  mutate(treatment = cut(year,breaks=c(0,2003,2020),
                     labels=c('pre-burn','post-burn'))) %>%
                  group_by(month, treatment) %>%
                  summarize(ndvi=mean(ndvi))

#View(ndvi_month_wide)

  p7 <- ggplot(ndvi_month_wide_burn, aes(x=month, y=ndvi, color=treatment)) +
      geom_point() +
      theme_few() +
      scale_color_few() +
       theme(legend.position=c(.6, .2)) 
  p7
    
```

In the burned plots, the greenest month is still August, even after the burn. 


### Question 5

What month is the snowiest on average?


#### Nicole's Code and Comments

```{r question 5}

ndsi_month_wide <-select(full_wide, -ndmi, -ndvi) %>%
                  group_by(month) %>%
                  summarize(ndsi=mean(ndsi))

#View(ndsi_month_wide)

  p8 <- ggplot(ndsi_month_wide, aes(x=month, y=ndsi)) +
      geom_point() +
      theme_few() +
      scale_color_few()
  p8
  
```

On average, the snowiest month is January, followed by February in a close second.

